---
title: "PA_Model"
author: "Bridget Sands"
date: "`r Sys.Date()`"
output: pdf_document
---

#### Bridget Sands
#### Harvard University
#### Applied Mathematics Senior Thesis
#### April 1st, 2024

# "PA_model.Rmd"

#### Notebook Purpose and Summary:
This notebook first combines the cleaned data for all tweleve seasons, casts and finalizes variable types, and fits the PA model and submodels.

#### Input:
Cleaned data for all tweleve seasons, after each has been run through the `Data_Cleaning_PA.ipynb` file.

#### Output:
For each model/submodel:
- `rds` data type--> R object of model
- `txt` of final model coefficients 

### Additional Notes:
Each model takes a significant amount of time to run. Some of them multiple days if fitted on the entire data. Beware.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Import necessary libraries
# data manipulation library
library(dplyr)
# model library
library(ordinal)
library(lme4)
library(ggeffects)
```


```{r}
# Read in initial data frame
df <- read.csv('PA_data/da14_PA.csv')
print(nrow(df))
```

```{r}
# Read in rest of data
files <- c('PA_data/da21_PA.csv', 'PA_data/da22_PA.csv', 'PA_data/da23_PA.csv', 'PA_data/ta14_PA.csv', 'PA_data/ta21_PA.csv', 'PA_data/ta22_PA.csv', 'PA_data/ta23_PA.csv', 'PA_data/mlb14_PA.csv', 'PA_data/mlb21_PA.csv', 'PA_data/mlb22_PA.csv', 'PA_data/mlb23_PA.csv')

for (file in files) {
  temp <- read.csv(file)
  # Append new file to old files
  df <- rbind(df, temp)
}
# Print length of full dataframe
print(nrow(df))
```

```{r}
# Check that condition treatments are correct
ls = c('AA', 'AAA', 'MLB')
yrs = c(2014, 2021, 2022, 2023)

for (l in ls){
  print(l)
  for (yr in yrs){
    print(yr)
    subset_df <- subset(df, (Year==yr) & (League == l), select = c(Pitch_timer, Bigger_bases, Defensive_shift_limits))
    cat('Unique Pitch Timer Values: ', unique(subset_df$Pitch_timer), '\n')
    cat('Unique Bigger Bases Values: ', unique(subset_df$Bigger_bases), '\n')
    cat('Unique Defensive Shift Limit Values: ', unique(subset_df$Defensive_shift_limits), '\n')
    cat('\n')
  }
}
```

```{r}
# Ensure the data does not include any instances where there are more than two strikes (not possible in reality)
df <- df[!(df$Strikes > 2), ]

# Ensure the data does not include any instances where there are more than three balls (not possible in reality)
df <- df[!(df$Balls > 3), ]

# Get rid of old useless column
df <- df[, -which(names(df) %in% c("X"))]
```

```{r}
# Cast Strikes, Balls, Outs as Integers
df$Strikes <- as.integer(df$Strikes)
df$Balls <- as.integer(df$Balls)
df$Outs <- as.integer(df$Outs)

# Ensure unique values are proper
cat("Unique Strikes values:", unique(df$Strikes), '\n')
cat("Unique Balls values:", unique(df$Balls), '\n')
cat("Unique Outs values:", unique(df$Outs), '\n')
```

```{r}
# Cast Batter_home, Bigger_bases, and ABS as logicals
df$Batter_home <- as.logical(df$Batter_home)
df$Bigger_bases <- as.logical(df$Bigger_bases)
df$ABS <- as.logical(df$ABS)

# Ensure unique values are proper
cat("Unique Batter_home values:", unique(df$Batter_home), '\n')
cat("Unique Bigger_bases values:", unique(df$Bigger_bases), '\n')
cat("Unique ABS values:", unique(df$ABS), '\n')
```

```{r}
# Cast Inning, Matchup_handed, Year, League as factors:
df$Inning <- as.factor(df$Inning)
df$Matchup_handed <- as.factor(df$Matchup_handed)
df$Year <- as.factor(df$Year)
df$League <- as.factor(df$League)

# Ensure unique values are proper:
cat("Unique Inning values:", unique(df$Inning), '\n')
cat("Unique Matchup_handed values:", unique(df$Matchup_handed), '\n')
cat("Unique Year values:", unique(df$Year), '\n')
cat("Unique League values:", unique(df$League), '\n')
```
```{r}
# Cast Pitch_timer and Defensive_shift_limits as factors
df$Pitch_timer <- as.factor(df$Pitch_timer)
df$Defensive_shift_limits <- as.factor(df$Defensive_shift_limits)

# Ensure unique values are proper:
cat("Unique Pitch_timer values:", unique(df$Pitch_timer), '\n')
cat("Unique Defensive_shift_limits values:", unique(df$Defensive_shift_limits), '\n')

# Cast Batter_id and Pitcher_id as factors
df$Batter_id <- as.factor(df$Batter_id)
df$Pitcher_id<- as.factor(df$Pitcher_id)

# Note that beacause of the magnitude of Batter_id and Pitcher_id  it does not make sense inspect
```

```{r}
# Set order for base states
men_ob_order <- c("Empty", "Men_On", "RISP", "Loaded")

# Cast Men_OB as an ordinal factor
df <- df %>%
  mutate(Men_OB = factor(Men_OB, ordered = TRUE, levels = men_ob_order))

# Ensure unique values are proper:
cat("Unique Men_OB values:", unique(df$Men_OB))
```


```{r}
# Set order for response variable
pitch_outcome_order <- c("Strike", "InPlay_Out", "Foul", "Ball", "Single", "Double", "Triple", "HR")

# Cast Pitch_outcome as an ordinal factor
df <- df %>%
  mutate(Pitch_outcome = factor(Pitch_outcome, ordered = TRUE, levels = pitch_outcome_order))

# Ensure unique values are proper:
cat("Unique Pitch_outcome values:", unique(df$Pitch_outcome))
```

```{r}
# Confirm final data types of each variable
str(df)
```

### Isolate the sub-count situations
```{r}
# Print number of rows in whole dataframe
nrow(df)
# Isolate full counts, print number of rows in the subset
df_fullCount <- subset(df, Balls==3&Strikes==2)
nrow(df_fullCount)
# Isolate empty counts, print number of rows in the subset
df_emptyCount <- subset(df, Balls==0&Strikes==0)
nrow(df_emptyCount)
# Isolate pitcher's counts, print number of rows in the subset
df_pitchersCount <- subset(df, Balls==0&Strikes==2)
nrow(df_pitchersCount)
# Isolate batter's counts, print number of rows in the subset
df_battersCount <- subset(df, Balls==3&Strikes==0)
nrow(df_battersCount)
```

### Regular PA Model:
```{r}
# Define control parameters
my_control <- clmm.control(trace = 1, method = "nlminb", abs.tol = 1, maxIter = 20, maxLineIter = 20, gradTol = .1, useMatrix = FALSE)

start_time <- Sys.time()

model <- clmm(Pitch_outcome ~ Strikes + Balls + Outs + Men_OB + Batter_home + Matchup_handed + Inning + Year + League + ABS + Pitch_timer + Bigger_bases + Defensive_shift_limits + (1|Batter_id) + (1|Pitcher_id), 
                data = df,
                link = "logit",
                control = my_control)

end_time <- Sys.time()
computation_time <- end_time - start_time
computation_time
# Export model
saveRDS(model, file = 'PA.rds')

s <- summary(model)
s
capture.output(s, file = "PA_OG.txt")

s2 <- fitted(model)
write.csv(as.data.frame(s2), file = "PA_OG_fitted.csv", row.names = FALSE)

print('Done with Regular Model, onto Full Count.')
```

#### Define formula to be called in other models:
```{r}
form <- Pitch_outcome ~ Strikes + Balls + Outs + Men_OB + Batter_home + Matchup_handed + Inning + Year + League + ABS + Pitch_timer + Bigger_bases + Defensive_shift_limits + (1|Batter_id) + (1|Pitcher_id)
```

### Full Count PA Model:
```{r}
start_time <- Sys.time()

model_fullCount <- clmm(form,
                        data = df_fullCount,
                        link = "logit",
                        control = my_control)

end_time <- Sys.time()
computation_time <- end_time - start_time
computation_time

saveRDS(model_fullCount, file = 'PA_fullCount.rds')

s_fullCount <- summary(model_fullCount)
capture.output(s_fullCount, file = "PA_fullCount.txt")

s2_fullCount <- fitted(model_fullCount)
write.csv(as.data.frame(s2_fullCount), file = "PA_fullCount_fitted.csv", row.names = FALSE)

print('Done with Full Count Model, onto Empty Count.')
```

### Empty Count PA Model
```{r}
start_time <- Sys.time()

model_emptyCount <- clmm(form, 
                        data = df_emptyCount,
                        link = "logit",
                        control = my_control)

end_time <- Sys.time()
computation_time <- end_time - start_time
computation_time

saveRDS(model_emptyCount, file = 'PA_emptyCount.rds')

s_emptyCount <- summary(model_emptyCount)
capture.output(s_emptyCount, file = "PA_emptyCount.txt")

s2_emptyCount <- fitted(model_emptyCount)
write.csv(as.data.frame(s2_emptyCount), file = "PA_emptyCount_fitted.csv", row.names = FALSE)

print('Done with Empty Count Model, onto Pitchers Count.')
```

### Pitcher's Count PA Model:
```{r}
start_time <- Sys.time()

model_pitchersCount <- clmm(form, 
                        data = df_pitchersCount,
                        link = "logit",
                        control = my_control)

end_time <- Sys.time()
computation_time <- end_time - start_time
computation_time

saveRDS(model_pitchersCount, file = 'PA_pitchersCount.rds')

s_pitchersCount <- summary(model_pitchersCount)
capture.output(s_pitchersCount, file = "PA_pitchersCount.txt")

s2_pitchersCount <- fitted(model_pitchersCount)
write.csv(as.data.frame(s2_pitchersCount), file = "PA_pitchersCount_fitted.csv", row.names = FALSE)

print('Done with Pitchers Count Model, onto Batters Count.')
```

### Batter's Count PA Model:
```{r}
start_time <- Sys.time()

model_battersCount <- clmm(form, 
                        data = df_battersCount,
                        link = "logit",
                        control = my_control)

end_time <- Sys.time()
computation_time <- end_time - start_time
computation_time

saveRDS(model_battersCount, file = 'PA_battersCount.rds')

s_battersCount <- summary(model_battersCount)
capture.output(s_battersCount, file = "PA_battersCount.txt")

s2_battersCount <- fitted(model_battersCount)
write.csv(as.data.frame(s2_battersCount), file = "PA_battersCount_fitted.csv", row.names = FALSE)

print('DONE WITH ALL!!!')
```
