---
title: "Acquire game_pks for 2022 ABS games in Triple-A Baseball:"
author: "Bridget Sands"
output: pdf_document
date: "2024-04-01"
---

#### Bridget Sands
#### Harvard University
#### Applied Mathematics Senior Thesis
#### April 1st, 2024

# "ABS_teams.Rmd"

#### Notebook Purpose and Summary:
This notebook is only used acquire and export the game_pks in which the automatic ball and strike system (ABS) was used in Triple-A baseball in its 2022 season.

#### Input:
Nothing specific from the user, input already set.

#### Output:
`csv` of game_pk values in which ABS was used from the 2022 Triple-A baseball season.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Import necessary libraries
library(dplyr)
library(ggeffects)
library(baseballr)
```


```{r}
# Get 2022 Triple-A Schedule
ta_22_sch <- mlb_schedule(season = 2022, level_ids = "11")

# Filter to only include regular season games
ta_22_sch <- ta_22_sch[ta_22_sch$series_description=='Regular Season', ]

# Cast the date as a date object
ta_22_sch$date <- as.Date(ta_22_sch$date)

# Inpsect data frame head
head(ta_22_sch)
```

```{r}
# Sort and view team options to identify proper spelling of names
sort(unique(ta_22_sch$teams_home_team_name))
```

```{r}
# Isolate the Charlotte Knights, spelled above 
char <- "Charlotte Knights"

# Isolate the names of the Pacific Coast League Teams
problems <- c("Albuquerque Isotopes", "El Paso Chihuahuas", "Las Vegas Aviators", "Oklahoma City Dodgers", "Reno Aces", "Round Rock Express", "Sacramento River Cats", "Salt Lake Bees", "Sugar Land Space Cowboys", "Tacoma Rainiers")
```

```{r}
# Isolate all game_pk values from games of the sesaon that the Charlotte Knights hosted
pks1 <- ta_22_sch$game_pk[ta_22_sch$teams_home_team_name == char]

# Isolate all game_pk values from games of the season on or after May 17th that the Pacific Coast Teams hosted
pks2 <- ta_22_sch$game_pk[ta_22_sch$teams_home_team_name %in% problems & ta_22_sch$date >= '2022-05-17']
```

```{r}
# Concatenate the list of the problematic pks
pks <- c(pks1, pks2)

# View 
pks
```

```{r}
# Convert list into data.frame
my_df <- data.frame(value = pks)

# Inspect
my_df
```
```{r}
# Write list to csv
write.csv(my_df, file = "problem_pks.csv", row.names = FALSE)
```







