---
title: "Data Acquisition from baseballR"
author: "Bridget Sands"
output: pdf_document
date: "2024-04-01"
---

#### Bridget Sands
#### Harvard University
#### Applied Mathematics Senior Thesis
#### April 1st, 2024

# "Data_Acquistion.Rmd"

### Note: This is the 1st and only notebook used for data acquisition. It is the first of all of the code files that should be used, because without it there is no data to import into the others.

#### Notebook Purpose and Summary:
This Rmarkdown notebook makes calls to baseballR's functions in order to acquire pitch by pitch data for use.

#### Input:
Needs user imput relative to year and league to call functions for.

#### Output:
`csv` season of data for the specified year and league.

### Additional Notes:
This file, regardless of the season/year pairing, takes a signficant amount of time to run. Beware.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Import necessary libraries
# baseballR, for access to data
require(baseballr)
# tidyverse, for data manipulation
require(tidyverse)
```

```{r}
# Initialize league id parameters to use
league_ids <- c(
  "MLB"="1", 
  "AAA"="11", 
  "AA"="12"
)

# Initialize year parameters to use
years <- c(2014, 2021, 2022, 2023)
```


### REMEMBER: Change Year and League
```{r}
# Distinguish the year and league to use for this iteration
league <- 'AA'
league_id <- league_ids[league]
year <- years[1]

paste("Getting data for", league, "for the", year, "season.")
```


```{r}
# Get MLB schedule for specific league and season
ids <- mlb_schedule(season = year, level_ids = league_id)

# Filter for only regular season games
ids <- filter(ids, series_description == "Regular Season")

# Inspect head of returned data
head(ids)

# Isolate the game_pks of data
pks <- ids$game_pk
```
```{r}
# Initialize pbp data by calling it for first game in pks list
pbp_data <- mlb_pbp(pks[1])

# Inspect
head(pbp_data)
```
```{r}
# Print length of pks, so we know how many we are iterating through
length(pks)

# Print pks to see
pks
```

```{r}
errors <- 0

# Call function and append data for all subsequent games in pks list
for (i in 2:length(pks)){
  # Implement a tryCatch to continue iteration if function returns nothing
  tryCatch({
    pbp_data <- rbind(pbp_data, mlb_pbp(pks[i]), fill=TRUE)
    
    # Print if i is divisable by 100, for progress tracking purposes
    if (i %% 100 == 0) {
      print(i)
    }
  }, error = function(e) {
    # Handle the error here, for example, print an error message
    errors <- errors + 1
    cat("Error occurred at iteration", i, ":", conditionMessage(e), "\n")
  })
}

paste("There were a total of", errors, "games that didn't have availble pbp data.")
```

### REMEMBER: Change file output
```{r}
#apply(mlb_pbp_23, 2, class)
write.csv(pbp, file = "dA14_r.csv")
```

