---
title: "SB_model"
author: "Bridget Sands"
date: "2024-04-01"
output: pdf_document
---

#### Bridget Sands
#### Harvard University
#### Applied Mathematics Senior Thesis
#### April 1st, 2024

# "SB_model.Rmd"

#### Notebook Purpose and Summary:
This notebook first combines the cleaned data for all tweleve seasons, casts and finalizes variable types, and fits the SB model and submodels.

#### Input:
Cleaned data for all tweleve seasons, after each has been run through the `Data_Cleaning_SB.ipynb` file.

#### Output:
For each model/submodel:
- `rds` data type--> R objcet of model
- `txt` of final model coefficients 

### Additional Notes:
Each model takes a significant amount of time to run. Its not as long as the PA_model(s), but it is still long.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Import necessary libraries
library(dplyr)
library(lme4)
# model library
library(ordinal)
library(lme4)
```

```{r}
# Read in initial data frame
df <- read.csv('SB_data/da14_SB.csv')
print(nrow(df))
```

```{r}
# Read in rest of data
files <- c('SB_data/da21_SB.csv', 'SB_data/da22_SB.csv', 'SB_data/da23_SB.csv', 'SB_data/ta14_SB.csv', 'SB_data/ta21_SB.csv', 'SB_data/ta22_SB.csv', 'SB_data/ta23_SB.csv', 'SB_data/mlb14_SB.csv', 'SB_data/mlb21_SB.csv', 'SB_data/mlb22_SB.csv', 'SB_data/mlb23_SB.csv')

for (file in files) {
  temp <- read.csv(file)
  # Append new file to old files
  df <- rbind(df, temp)
}
# Print length of full dataframe
print(nrow(df))
```

```{r}
# Print number of missing values in firstID
cat('Number of nas in firstID: ', sum(is.na(df$firstID)), '\n')
# Print number of missing values in Batter_id
cat('Number of nas in Batter_id: ', sum(is.na(df$Batter_id)), '\n')

# Filter data so there are no instances of NAs in firstID
df <- df[!is.na(df$firstID), ]
```

```{r}
# Check that proper conditions hold
ls = c('AA', 'AAA', 'MLB')
yrs = c(2014, 2021, 2022, 2023)

for (l in ls){
  print(l)
  for (yr in yrs){
    print(yr)
    subset_df <- subset(df, (Year==yr) & (League == l), select = c(Pitch_timer, Bigger_bases, Defensive_shift_limits))
    cat('Unique Pitch Timer Values: ', unique(subset_df$Pitch_timer), '\n')
    cat('Unique Bigger Bases Values: ', unique(subset_df$Bigger_bases), '\n')
    cat('Unique Defensive Shift Limit Values: ', unique(subset_df$Defensive_shift_limits), '\n')
    cat('\n')
  }
}
```

```{r}
# Inspect values of Strikes, Balls, and Outs
table(df$Strikes)
table(df$Balls)
table(df$Outs)
```

```{r}
# Ensure the data does not include any instances where there are more than two strikes (not possible in reality)
df <- df[!(df$Strikes > 2), ]

# Ensure the data does not include any instances where there are more than three balls (not possible in reality)
df <- df[!(df$Balls > 3), ]

# Get rid of old useless column
df <- df[, -which(names(df) %in% c("X"))]
```

```{r}
# Cast Strikes, Balls, Outs as Integers
df$Strikes <- as.integer(df$Strikes)
df$Balls <- as.integer(df$Balls)
df$Outs <- as.integer(df$Outs)

# Ensure unique values are proper
cat("Unique Strikes values:", unique(df$Strikes), '\n')
cat("Unique Balls values:", unique(df$Balls), '\n')
cat("Unique Outs values:", unique(df$Outs), '\n')
```

```{r}
# Cast Batter_home, Bigger_bases, ABS, and SBA as logicals
df$Batter_home <- as.logical(df$Batter_home)
df$Bigger_bases <- as.logical(df$Bigger_bases)
df$ABS <- as.logical(df$ABS)
df$SBA <- as.logical(df$SBA)

# Ensure unique values are proper
cat("Unique Batter_home values:", unique(df$Batter_home), '\n')
cat("Unique Bigger_bases values:", unique(df$Bigger_bases), '\n')
cat("Unique ABS values:", unique(df$ABS), '\n')
cat("Unique SBA values:", unique(df$SBA), '\n')
```

```{r}
# Cast Inning, AB_pickoffs, Year, League as factors:
df$Inning <- as.factor(df$Inning)
df$AB_pickoffs <- as.factor(df$AB_pickoffs)
df$Year <- as.factor(df$Year)
df$League <- as.factor(df$League)

# Ensure unique values are proper:
cat("Unique Inning values:", unique(df$Inning), '\n')
cat("Unique AB_pickoffs values:", unique(df$AB_pickoffs), '\n')
cat("Unique Year values:", unique(df$Year), '\n')
cat("Unique League values:", unique(df$League), '\n')
```
```{r}
table(df$AB_pickoffs)
```

```{r}
# Cast Pitch_timer and Defensive_shift_limits as factors
df$Pitch_timer <- as.factor(df$Pitch_timer)
df$Defensive_shift_limits <- as.factor(df$Defensive_shift_limits)

# Ensure unique values are proper:
cat("Unique Pitch_timer values:", unique(df$Pitch_timer), '\n')
cat("Unique Defensive_shift_limits values:", unique(df$Defensive_shift_limits), '\n')

# Cast Batter_id, Pitcher_id, and firstID as factors
df$Batter_id <- as.factor(df$Batter_id)
df$Pitcher_id<- as.factor(df$Pitcher_id)
df$firstID<- as.factor(df$firstID)

# Note that beacause of the magnitude of Batter_id, Pitcher_id, and firstID, it does not make sense inspect
```

```{r}
# Confirm final data types of each variable
str(df)
```

### Regular SB Model:
```{r}
# Now, let's fit this model
start_time <- Sys.time()
model <- glmer(SBA ~ Strikes + Balls + Outs + Batter_home + Inning  + Year + League + AB_pickoffs + ABS + Pitch_timer + Bigger_bases + Defensive_shift_limits + (1|Batter_id) + (1|Pitcher_id) + (1|firstID), 
               verbose = 2,
               data = df, 
               family = binomial,
               control = glmerControl(calc.derivs = FALSE, tolPwrss = .1, optimizer="nloptwrap", optCtrl = list(algorithm = "NLOPT_LN_BOBYQA", ftol_abs = 1)))

end_time <- Sys.time()
computation_time <- end_time - start_time
computation_time

saveRDS(model, file = 'SB_F_315.rds')
s <- summary(model)
capture.output(s, file = "SB_315.txt")

print('Done with regular, onto without AB_pickoffs.')
```

### Regular SB Model, no AB_pickoffs:
```{r}
start_time <- Sys.time()
model_AB <- glmer(SBA ~ Strikes + Balls + Outs + Batter_home + Inning  + Year + League + ABS + Pitch_timer + Bigger_bases + Defensive_shift_limits + (1|Batter_id) + (1|Pitcher_id) + (1|firstID), 
               verbose = 2,
               data = df, 
               family = binomial,
               control = glmerControl(calc.derivs = FALSE, tolPwrss = .1, optimizer="nloptwrap", optCtrl = list(algorithm = "NLOPT_LN_BOBYQA", ftol_abs = 1)))

end_time <- Sys.time()
computation_time <- end_time - start_time
computation_time

saveRDS(model_AB, file = 'SB_F_315_NO_AB.rds')
s_AB <- summary(model_AB)
capture.output(s_AB, file = "SB_315_NO_AB.txt")

print('Done with regular without AB_pickoffs, onto full count.')
```

### Isolate the sub-count situations
```{r}
# Print number of rows in whole dataframe
nrow(df)
# Isolate full counts, print number of rows in the subset
df_fullCount <- subset(df, Balls==3&Strikes==2)
nrow(df_fullCount)
# Isolate empty counts, print number of rows in the subset
df_emptyCount <- subset(df, Balls==0&Strikes==0)
nrow(df_emptyCount)
# Isolate pitcher's counts, print number of rows in the subset
df_pitchersCount <- subset(df, Balls==0&Strikes==2)
nrow(df_pitchersCount)
# Isolate batter's counts, print number of rows in the subset
df_battersCount <- subset(df, Balls==3&Strikes==0)
nrow(df_battersCount)
```

### Full Count SB Model:
```{r}
# Now, let's fit this model
start_time <- Sys.time()
model_FC <- glmer(SBA ~ Strikes + Balls + Outs + Batter_home + Inning  + Year + League + AB_pickoffs + ABS + Pitch_timer + Bigger_bases + Defensive_shift_limits + (1|Batter_id) + (1|Pitcher_id) + (1|firstID), 
               verbose = 2,
               data = df_fullCount, 
               family = binomial,
               control = glmerControl(calc.derivs = FALSE, tolPwrss = .1, optimizer="nloptwrap", optCtrl = list(algorithm = "NLOPT_LN_BOBYQA", ftol_abs = 1)))

end_time <- Sys.time()
computation_time <- end_time - start_time
computation_time

saveRDS(model_FC, file = 'SB_F_315_fullCount.rds')
s_FC <- summary(model_FC)
capture.output(s_FC, file = "SB_315_fullCount.txt")

print('Done with regular FullCount, onto FullCount without AB_pickoffs.')
```

### Full Count SB Model, no AB_pickoffs:
```{r}
start_time <- Sys.time()
model_AB_FC <- glmer(SBA ~ Strikes + Balls + Outs + Batter_home + Inning  + Year + League + ABS + Pitch_timer + Bigger_bases + Defensive_shift_limits + (1|Batter_id) + (1|Pitcher_id) + (1|firstID), 
               verbose = 2,
               data = df_fullCount, 
               family = binomial,
               control = glmerControl(calc.derivs = FALSE, tolPwrss = .1, optimizer="nloptwrap", optCtrl = list(algorithm = "NLOPT_LN_BOBYQA", ftol_abs = 1)))

end_time <- Sys.time()
computation_time <- end_time - start_time
computation_time

saveRDS(model_AB_FC, file = 'SB_F_315_NO_AB_fullCount.rds')
s_AB_FC <- summary(model_AB_FC)
capture.output(s_AB_FC, file = "SB_315_NO_AB_fullCount.txt")

print('Done with full count without AB_pickoffs, onto empty count.')
```

### Empty Count SB Model:
```{r}
# Now, let's fit this model
start_time <- Sys.time()
model_EC <- glmer(SBA ~ Strikes + Balls + Outs + Batter_home + Inning  + Year + League + AB_pickoffs + ABS + Pitch_timer + Bigger_bases + Defensive_shift_limits + (1|Batter_id) + (1|Pitcher_id) + (1|firstID), 
               verbose = 2,
               data = df_emptyCount, 
               family = binomial,
               control = glmerControl(calc.derivs = FALSE, tolPwrss = .1, optimizer="nloptwrap", optCtrl = list(algorithm = "NLOPT_LN_BOBYQA", ftol_abs = 1)))

end_time <- Sys.time()
computation_time <- end_time - start_time
computation_time

saveRDS(model_EC, file = 'SB_F_315_emptyCount.rds')
s_EC <- summary(model_EC)
capture.output(s_EC, file = "SB_315_emptyCount.txt")

print('Done with regular, onto without AB_pickoffs.')
```

### Empty Count SB Model, no AB_pickoffs:
```{r}
start_time <- Sys.time()
model_AB_EC <- glmer(SBA ~ Strikes + Balls + Outs + Batter_home + Inning  + Year + League + ABS + Pitch_timer + Bigger_bases + Defensive_shift_limits + (1|Batter_id) + (1|Pitcher_id) + (1|firstID), 
               verbose = 2,
               data = df_emptyCount, 
               family = binomial,
               control = glmerControl(calc.derivs = FALSE, tolPwrss = .1, optimizer="nloptwrap", optCtrl = list(algorithm = "NLOPT_LN_BOBYQA", ftol_abs = 1)))

end_time <- Sys.time()
computation_time <- end_time - start_time
computation_time

saveRDS(model_AB_EC, file = 'SB_F_315_NO_AB_emptyCount.rds')
s_AB_EC <- summary(model_AB_EC)
capture.output(s_AB_EC, file = "SB_315_NO_AB_emptyCount.txt")

print('Done with empty count without AB_pickoffs, onto pitchers count.')
```

### Pitcher's Count SB Model:
```{r}
# Now, let's fit this model
start_time <- Sys.time()
model_PC <- glmer(SBA ~ Strikes + Balls + Outs + Batter_home + Inning  + Year + League + AB_pickoffs + ABS + Pitch_timer + Bigger_bases + Defensive_shift_limits + (1|Batter_id) + (1|Pitcher_id) + (1|firstID), 
               verbose = 2,
               data = df_pitchersCount, 
               family = binomial,
               control = glmerControl(calc.derivs = FALSE, tolPwrss = .1, optimizer="nloptwrap", optCtrl = list(algorithm = "NLOPT_LN_BOBYQA", ftol_abs = 1)))

end_time <- Sys.time()
computation_time <- end_time - start_time
computation_time

saveRDS(model_PC, file = 'SB_F_315_pitchersCount.rds')
s_PC <- summary(model_PC)
capture.output(s_PC, file = "SB_315_pitchersCount.txt")

print('Done with regular pitchers count, onto without AB_pickoffs.')
```

### Pitcher's Count SB Model, no AB_pickoffs:
```{r}
start_time <- Sys.time()
model_AB_PC <- glmer(SBA ~ Strikes + Balls + Outs + Batter_home + Inning  + Year + League + ABS + Pitch_timer + Bigger_bases + Defensive_shift_limits + (1|Batter_id) + (1|Pitcher_id) + (1|firstID), 
               verbose = 2,
               data = df_pitchersCount, 
               family = binomial,
               control = glmerControl(calc.derivs = FALSE, tolPwrss = .1, optimizer="nloptwrap", optCtrl = list(algorithm = "NLOPT_LN_BOBYQA", ftol_abs = 1)))

end_time <- Sys.time()
computation_time <- end_time - start_time
computation_time

saveRDS(model_AB_PC, file = 'SB_F_315_NO_AB_pitchersCount.rds')
s_AB_PC <- summary(model_AB_PC)
capture.output(s_AB_PC, file = "SB_315_NO_AB_pitchersCount.txt")

print('Done with pitchers count without AB_pickoffs, onto batters count.')
```

### Batter's Count SB Model:
```{r}
# Now, let's fit this model
start_time <- Sys.time()
model_BC <- glmer(SBA ~ Strikes + Balls + Outs + Batter_home + Inning  + Year + League + AB_pickoffs + ABS + Pitch_timer + Bigger_bases + Defensive_shift_limits + (1|Batter_id) + (1|Pitcher_id) + (1|firstID), 
               verbose = 2,
               data = df_battersCount, 
               family = binomial,
               control = glmerControl(calc.derivs = FALSE, tolPwrss = .1, optimizer="nloptwrap", optCtrl = list(algorithm = "NLOPT_LN_BOBYQA", ftol_abs = 1)))

end_time <- Sys.time()
computation_time <- end_time - start_time
computation_time

saveRDS(model_BC, file = 'SB_F_315_battersCount.rds')
s_BC <- summary(model_BC)
capture.output(s_BC, file = "SB_315_battersCount.txt")

print('Done with regular batters count, onto without AB_pickoffs.')
```

### Batter's Count SB Model, no AB_pickoffs:
```{r}
start_time <- Sys.time()
model_AB_BC <- glmer(SBA ~ Strikes + Balls + Outs + Batter_home + Inning  + Year + League + ABS + Pitch_timer + Bigger_bases + Defensive_shift_limits + (1|Batter_id) + (1|Pitcher_id) + (1|firstID), 
               verbose = 2,
               data = df_battersCount, 
               family = binomial,
               control = glmerControl(calc.derivs = FALSE, tolPwrss = .1, optimizer="nloptwrap", optCtrl = list(algorithm = "NLOPT_LN_BOBYQA", ftol_abs = 1)))

end_time <- Sys.time()
computation_time <- end_time - start_time
computation_time

saveRDS(model_AB_BC, file = 'SB_F_315_NO_AB_battersCount.rds')
s_AB_BC <- summary(model_AB_BC)
capture.output(s_AB_BC, file = "SB_315_NO_AB_battersCount.txt")

print('DONE WITH ALL!!!')
```

